Copulas and their potential for ecology

### 2019-10-11 Paper.Rmd
adding some stuff to Discussion

<!--***DAN: remianing to do:
DONE 1) Edit abstract
2) Go through the whole thing one more time
3) send it out to coauthors (done) and maybe get feedback
4)make sure to say something about tropical systems
DONE 5) maybe write to the editors saying this is coming soon?
6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
DONE 7) change the box around the Cedar Creek data in the main fig
8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
9) maybe go back and read the Ecol Monogr reviews again
-->

@@

<!--Partial Person correlation-->
We carried out most of our analyses on data ranks, $u_i$ and $v_i$, for good reasons 
int he Inrtoduction, Background, and methods sections and 
reviewed here; but under some circumstances it may also be appropriate to use 
techniques which parallel our nonparametric statistics but that use unranked data, 
$x_i$, $y_i$. As mentioned previously, @Genest2007 and others
recommend carrying out inferences about dependence structures (which was our
goal) using normalized ranks, stating explicitly "statistical inference concerning
dependence structures should always be based on ranks." The reason for this is
that measures of dependence which use unranked data conflate information about
the maginal distributions of variables with information about the dependence
between variables. To provide an ecological example, suppose the true population densities $p_{1,t}$
and $p_{2,t}$ of two species of fish in a lake in year $t$ are unknown,
but are assessed using catch per unit effort (CPUE), a standard approach in fisheries
science. The CPUE measurements, which we denote $c_{1,t}$ and $c_{2,t}$ may be more
or less correlated through time than the true densities would be, if they were known,
if the function $f$ relating $p_{i,t}$ and $c_{i,t}$ is nonlinear and if Pearson
correlation is used. The function $f$ modifies the marginal distributions, which modifies
measures such as Pearson correlation that conflate dependence and marginal information.
Rank-based measures such as Spearman or Kendall correlation, or any of our copula approaches,
will be unaffected by nonlinear functions such as $f$. The same difficulty will pertain in any
case for which a measurement of an ecological quantity is a nonlinear index of the true quantity
of interest. Nevertheless, if one is certain that measurements are linearly related to 
the true quantities being measured, statistics based on unranked data may be more appropriate 
under some circumstances, in part because ecological mechanisms which produce relationships 
between variables are influenced not only by the ordinal informatio in ranks but also 
by the cardinal information in the unranked data. Judgements of what statistics
are appropriate depend, of course, on the purpose of the analysis; the goal of this 
study was to study dependence in isolation, so we used ranks. If one is interested
in tail associations using cardinal data, a partial Pearson correlation can developed
straightforwardly by replacing $u$ and $v$ in equation \ref{eq:partialspearman}
by the $x$ and $y$, but still computing the sum in that equation over data points
with ranks, $u_i$ and $v_i$ constrained by the bounds $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$. This would result in a different approach to
tail associations that would conflate dependence and marginal information but 
that may also be useful for understanding ecological relationships.

<!--6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.-->
Copula methods are numerous, and go well beyond the cases we have considered
in this introductory paper; we hope our work helps inspire applications of copulas
in ecology beyond the specific tools we have employed. For instance, multivariate copulas are useful for 
studying relationships between multiple interacting variables, and have been used in finance
to model tail risk for portfolio optimization problems [@Czado2019; @Joe2011]. 
Analogous ecological applications
exist that may be amenable to the same approach, e.g., ecosystem functioning variables
such as total primary productivity or carbon flux are the sum of the contributions of
multiple species in the same way that athe value of an investment portfolio is the sum 
of the values of its constituent assets. Ecologists may likewise be interested in
managing the risk that an ecosystem functioning variable will take an extreme value. 
As another example, we assumed for this paper that univariate marginal 
distributions have continuous, strictly 
monotonic distribution functions; this was for simplicity and because such a simpler
approach was sufficient for our research questions. But count data in ecology have been analyzed
with approaches not making this assumption [@Anderson2018].

>>>>

<!--***DAN: remianing to do:
DONE 1) Edit abstract
2) Go through the whole thing one more time
3) send it out to coauthors (done) and maybe get feedback, ESP FROM lAWRENCE
4)make sure to say something about tropical systems
DONE 5) maybe write to the editors saying this is coming soon?
DONE 6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
7) change the box around the Cedar Creek data in the main fig
DONE 8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
9) maybe go back and read the Ecol Monogr reviews again
-->

@@

<!--Partial Person correlation-->
We carried out most of our analyses on data ranks, $u_i$ and $v_i$, for good reasons 
mentioned in the Inrtoduction, Background, and methods sections and 
reviewed here; but under some circumstances it may also be appropriate to use 
techniques which parallel our nonparametric statistics but that use unranked data, 
$x_i$, $y_i$. As mentioned previously, @Genest2007 and others
recommend carrying out inferences about dependence structures (which was our
goal) using normalized ranks, stating explicitly "statistical inference concerning
dependence structures should always be based on ranks." The reason for this is
that measures of dependence which use unranked data conflate information about
the maginal distributions of variables with information about the dependence
between variables. To provide an ecological example, suppose the true population densities $p_{1,t}$
and $p_{2,t}$ of two species of fish in a lake in year $t$ are unknown,
but are assessed using catch per unit effort (CPUE), a standard approach in fisheries
science. The CPUE measurements, which we denote $c_{1,t}$ and $c_{2,t}$ may be more
or less correlated through time than the true densities would be, if they were known,
if the function $f$ relating $p_{i,t}$ and $c_{i,t}$ is nonlinear and if Pearson
correlation is used. The function $f$ modifies the marginal distributions, which modifies
measures such as Pearson correlation that conflate dependence and marginal information.
Rank-based measures such as Spearman or Kendall correlation, or any of our copula approaches,
will be unaffected by nonlinear monotonic functions such as $f$. The same difficulty will pertain in any
case for which a measurement of an ecological quantity is a nonlinear index of the true quantity
of interest. Nevertheless, if one is certain that measurements are linearly related to 
the true quantities being measured, statistics based on unranked data may be more appropriate 
under some circumstances, in part because ecological mechanisms which produce relationships 
between variables are influenced not only by the ordinal informatio in ranks but also 
by the cardinal information in the unranked data. Judgements of what statistics
are appropriate depend, of course, on the purpose of the analysis; the goal of this 
study was to study dependence in isolation, so we used ranks. If one is interested
in tail associations using cardinal data, a partial Pearson correlation can developed
straightforwardly by replacing $u$ and $v$ in equation \ref{eq:partialspearman}
by the $x$ and $y$, but still computing the sum in that equation over data points
with ranks, $u_i$ and $v_i$ constrained by the bounds $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$. This would result in a different approach to
tail associations that would conflate dependence and marginal information but 
that may also be useful for understanding ecological relationships.

<!--additional copula methods-->
Copula methods are numerous, and go well beyond the cases we have considered
in this introductory paper; we hope our work helps inspire applications of copulas
in ecology both of and 
beyond the specific tools we have employed. For instance, multivariate copulas are useful for 
studying relationships between multiple interacting variables, and have been used in finance
to model tail risk for portfolio optimization problems [@Czado2019; @Joe2011]. 
Analogous ecological applications
exist that may be amenable to the same approach, e.g., ecosystem functioning variables
such as total primary productivity or carbon flux are the sum of the contributions of
multiple species in the same way that the value of an investment portfolio is the sum 
of the values of its constituent assets. Ecologists may likewise be interested in
managing the risk that an ecosystem functioning variable will take an extreme value. 
As another example, we assumed for this paper that univariate marginal 
distributions have continuous, strictly 
monotonic distribution functions; this was for simplicity and because the simpler
approach was sufficient for our research questions. But count data in ecology have been analyzed
with approaches not making this assumption [@Anderson2018].
